---
title: "Silicon Benchmark Forecast Diagnostics"
subtitle: "Team 16 — internal interpretation report"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    self_contained: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center",
  dpi = 120
)

suppressPackageStartupMessages({
  library(nalanda)
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(scales)
  library(tidyr)
})

root_candidates <- unique(c(
  normalizePath(".", winslash = "/", mustWork = FALSE),
  normalizePath("..", winslash = "/", mustWork = FALSE)
))
is_project_root <- vapply(
  root_candidates,
  function(x) {
    file.exists(file.path(x, "inputs", "tier3_outcomes.csv")) &&
      file.exists(file.path(
        x, "outputs", "derived", "nalanda_forecast_aggregation_long.csv"
      ))
  },
  logical(1)
)

if (!any(is_project_root)) {
  stop("Could not locate the Silicon Benchmark project root.")
}

project_root <- root_candidates[which(is_project_root)[1]]
project_file <- function(...) file.path(project_root, ...)

outcomes <- read_csv(
  project_file("inputs", "tier3_outcomes.csv"),
  show_col_types = FALSE
)
interventions <- read_csv(
  project_file("inputs", "tier3_interventions.csv"),
  show_col_types = FALSE
)
models_config <- read_csv(
  project_file("config", "models.csv"),
  show_col_types = FALSE
)
forecast_stages <- read_csv(
  project_file(
    "outputs", "derived", "nalanda_forecast_aggregation_long.csv"
  ),
  show_col_types = FALSE
)
prompt_level <- forecast_stages |>
  filter(aggregation_level == "prompt") |>
  transmute(
    condition,
    outcome,
    forecast_model = model_id,
    prompt_variant = prompt_id,
    prompt_ate = estimate
  )
model_level <- forecast_stages |>
  filter(aggregation_level == "model") |>
  transmute(
    condition,
    outcome,
    forecast_model = model_id,
    model_ate = estimate
  )
candidate <- read_csv(
  project_file(
    "outputs", "derived", "team_16_T3_candidate_median_unscaled.csv"
  ),
  show_col_types = FALSE
)
all_model_candidate <- read_csv(
  project_file(
    "outputs", "derived",
    "team_16_T3_diagnostic_all_models_median_unscaled.csv"
  ),
  show_col_types = FALSE
)
primary_roster <- read_csv(
  project_file("outputs", "derived", "primary_model_roster.csv"),
  show_col_types = FALSE
)

aggregation_comparison_path <- project_file(
  "outputs", "derived", "aggregation_method_comparison.csv"
)
mean_sensitivity_available <- file.exists(aggregation_comparison_path)
if (mean_sensitivity_available) {
  aggregation_comparison <- read_csv(
    aggregation_comparison_path,
    show_col_types = FALSE
  )
}

prompt_dispersion <- read_csv(
  project_file(
    "outputs", "diagnostics", "prompt_disagreement_by_cell.csv"
  ),
  show_col_types = FALSE
) |>
  transmute(
    condition,
    outcome,
    forecast_model = model_id,
    n_prompts = n_contributors,
    prompt_min_fraction = min / scale_width,
    prompt_max_fraction = max / scale_width,
    prompt_range_fraction = range_per_scale
  )

outcome_order <- outcomes$outcome
condition_order <- sort(unique(interventions$condition))
model_names <- sort(unique(model_level$forecast_model))
prompt_names <- sort(unique(prompt_level$prompt_variant))
n_models <- length(model_names)
n_prompts <- length(prompt_names)
n_cells <- nrow(candidate)

represented_models <- models_config |>
  filter(model_id %in% model_names) |>
  arrange(family, model_id)
n_families <- n_distinct(represented_models$family)
all_family_composition <- represented_models |>
  group_by(family) |>
  summarize(
    n_models = n(),
    models = paste(model_id, collapse = ", "),
    .groups = "drop"
  )
primary_models <- models_config |>
  filter(model_id %in% primary_roster$model_id) |>
  arrange(family, model_id)
n_primary_families <- n_distinct(primary_models$family)
primary_family_composition <- primary_models |>
  group_by(family) |>
  summarize(
    n_models = n(),
    models = paste(model_id, collapse = ", "),
    .groups = "drop"
  )

candidate_enriched <- candidate |>
  left_join(
    select(outcomes, outcome, native_min, native_max, definition),
    by = "outcome"
  ) |>
  mutate(
    scale_width = native_max - native_min,
    ate_fraction = ate / scale_width,
    outcome = factor(outcome, levels = outcome_order),
    condition = factor(condition, levels = rev(condition_order))
  )
all_model_candidate_enriched <- all_model_candidate |>
  left_join(
    select(outcomes, outcome, native_min, native_max, definition),
    by = "outcome"
  ) |>
  mutate(
    scale_width = native_max - native_min,
    ate_fraction = ate / scale_width,
    outcome = factor(outcome, levels = outcome_order),
    condition = factor(condition, levels = rev(condition_order))
  )
candidate_fill_limit <- max(abs(c(
  candidate_enriched$ate_fraction,
  all_model_candidate_enriched$ate_fraction
)), na.rm = TRUE)
roster_sensitivity <- candidate |>
  rename(primary_ate = ate) |>
  inner_join(
    rename(all_model_candidate, all_models_ate = ate),
    by = c("condition", "outcome")
  ) |>
  left_join(
    transmute(outcomes, outcome, scale_width = native_max - native_min),
    by = "outcome"
  ) |>
  mutate(
    absolute_gap_fraction = abs(primary_ate - all_models_ate) / scale_width
  )

prompt_enriched <- prompt_level |>
  left_join(
    select(outcomes, outcome, native_min, native_max),
    by = "outcome"
  ) |>
  mutate(
    scale_width = native_max - native_min,
    prompt_ate_fraction = prompt_ate / scale_width
  )

median_abs_effect <- median(abs(candidate_enriched$ate_fraction))
max_abs_effect <- max(abs(candidate_enriched$ate_fraction))
median_prompt_gap <- median(prompt_dispersion$prompt_range_fraction)
max_prompt_gap <- max(prompt_dispersion$prompt_range_fraction)
positive_cells <- sum(candidate_enriched$ate > 0)
negative_cells <- sum(candidate_enriched$ate < 0)
zero_cells <- sum(candidate_enriched$ate == 0)

kable_simple <- function(x, digits = 3, caption = NULL) {
  knitr::kable(x, digits = digits, caption = caption)
}
```

# Executive summary

This report summarizes **internal forecast behavior**, not predictive accuracy.
The benchmark outcomes remain unknown, so correlations and agreement can reveal
stability or shared assumptions but cannot establish that a model is correct.

```{r executive-callout, results='asis'}
if (n_models == 1L) {
  cat(
    "> **Current limitation:** The present files contain one model (`",
    model_names,
    "`) and ", n_prompts, " prompt variants. The candidate file is therefore ",
    "a within-model prompt ensemble, not yet a multi-model consensus.\n\n",
    sep = ""
  )
} else {
  cat(
    "> The present files contain **", n_models, " models** and **", n_prompts,
    " prompt variants**. Model-level values are aggregated before the models ",
    "receive equal weight.\n\n",
    sep = ""
  )
}
```

- Coverage: **`r n_cells` of 208 intervention–outcome cells**.
- Models represented: **`r n_models`**.
- Fixed prompt variants represented: **`r n_prompts`**.
- Median absolute candidate effect: **`r percent(median_abs_effect, accuracy = 0.01)` of the outcome scale**.
- Largest absolute candidate effect: **`r percent(max_abs_effect, accuracy = 0.01)` of the outcome scale**.
- Directional balance: **`r positive_cells` positive**, **`r negative_cells` negative**, and **`r zero_cells` exactly zero** forecasts.
- Median prompt-to-prompt range: **`r percent(median_prompt_gap, accuracy = 0.01)` of the outcome scale**; maximum: **`r percent(max_prompt_gap, accuracy = 0.01)`**.

# Run contents and completeness

```{r run-summary}
run_summary <- tibble::tibble(
  Item = c(
    "Models in derived files",
    "Prompt variants",
    "Interventions",
    "Outcomes",
    "Unique intervention-outcome cells",
    "Expected intervention-outcome cells",
    "Missing numeric forecasts"
  ),
  Value = c(
    n_models,
    n_prompts,
    n_distinct(candidate$condition),
    n_distinct(candidate$outcome),
    n_cells,
    16L * 13L,
    sum(is.na(forecast_stages$estimate))
  )
)
kable_simple(run_summary)
```

```{r model-settings}
settings_used <- models_config |>
  filter(model_id %in% model_names) |>
  select(
    model_id,
    model,
    family,
    temperature,
    output_mode,
    n_completions,
    roster_status
  )
kable_simple(settings_used, caption = "Configured settings for models represented in the report")
```

# Candidate consensus forecast map

This figure shows the primary provisional consensus from the three preselected
family representatives. For every intervention and outcome, repeated
completions are combined within prompts, prompts within models, and finally the
`r n_primary_families` represented model families receive equal weight. Medians
are used at every step to reduce the influence of an unusually extreme forecast.

<div class="alert alert-info">
<strong>What is an average treatment effect (ATE)?</strong><br>
An ATE is the predicted average outcome in the intervention group minus the
predicted average in the control group. For example, this study includes an
actual <code>donation_ams</code> outcome ranging from $0 to $10. An ATE of $1
would mean that participants receiving the intervention are forecast to donate
one dollar more, on average, than participants reading a neutral control text.
On the normalized heatmap that equals 10% of the donation scale. Similarly, a
5-point ATE on a 0–100 outcome equals 5% of its scale, while an ATE of 0.05 for
the binary newsletter-signup outcome means a five-percentage-point difference.
The sign matters: a negative ATE means the intervention group is forecast to
score lower than the control group, which can be desirable for an outcome such
as distrust.
</div>

```{r candidate-composition}
kable_simple(
  primary_family_composition,
  caption = "Preselected models contributing to the primary provisional consensus"
)
```

Each tile expresses the consensus ATE as a fraction of that outcome's possible
range. The map is useful for seeing direction and relative magnitude across the
complete intervention-by-outcome grid, spotting unexpectedly extreme cells,
and checking whether substantive patterns are coherent. It does not show
predictive accuracy because the human benchmark results remain hidden.

```{r candidate-heatmap, fig.height=8}
ggplot(candidate_enriched, aes(x = outcome, y = condition, fill = ate_fraction)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-candidate_fill_limit, candidate_fill_limit),
    labels = label_percent(accuracy = 0.1),
    name = "ATE as %\nof scale"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Primary consensus: three preselected family representatives"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

## All-model diagnostic sensitivity map

This second map repeats the same hierarchical median rule using all `r n_models`
models with complete full-run results. Models are still combined within their
developer families before the `r n_families` families receive equal weight, so
a family does not gain extra final weight merely because more variants were
tested. This map is diagnostic only: it asks whether broadening the model roster
materially changes the pattern of forecasts. Both maps use the same color scale,
so their colors can be compared directly.

```{r all-model-composition}
kable_simple(
  all_family_composition,
  caption = "All completed models contributing to the diagnostic sensitivity map"
)
```

```{r all-model-heatmap, fig.height=8}
ggplot(
  all_model_candidate_enriched,
  aes(x = outcome, y = condition, fill = ate_fraction)
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-candidate_fill_limit, candidate_fill_limit),
    labels = label_percent(accuracy = 0.1),
    name = "ATE as %\nof scale"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Diagnostic sensitivity: all completed models"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

Across the 208 cells, the median absolute difference between the primary and
all-model consensuses is
**`r percent(median(roster_sensitivity$absolute_gap_fraction), accuracy = 0.01)`
of the outcome scale**; the largest difference is
**`r percent(max(roster_sensitivity$absolute_gap_fraction), accuracy = 0.01)`**.

## Largest positive and negative forecasts

These are useful for checking face validity and scale interpretation. They are
not evidence that the corresponding interventions will produce the largest
human effects.

```{r extreme-tables}
extremes <- candidate_enriched |>
  mutate(
    direction = if_else(ate_fraction >= 0, "Positive", "Negative"),
    scale_percent = 100 * ate_fraction
  ) |>
  group_by(direction) |>
  slice_max(abs(ate_fraction), n = 8, with_ties = FALSE) |>
  ungroup() |>
  arrange(direction, desc(abs(ate_fraction))) |>
  select(direction, condition, outcome, ate, scale_percent)

kable_simple(
  extremes,
  digits = 2,
  caption = "Largest absolute candidate effects after normalizing by outcome range"
)
```

# Outcome-level calibration

This figure asks whether the forecasting ensemble systematically assigns larger
effects to some outcomes than others after accounting for their different native
scales. Outcomes are sorted by the median absolute consensus effect, with the
largest typical effects at the top.

```{r outcome-calibration}
outcome_summary <- candidate_enriched |>
  group_by(outcome) |>
  summarize(
    median_ate = median(ate),
    median_absolute_scale_percent = 100 * median(abs(ate_fraction)),
    maximum_absolute_scale_percent = 100 * max(abs(ate_fraction)),
    .groups = "drop"
  ) |>
  mutate(outcome = reorder(outcome, median_absolute_scale_percent))

ggplot(
  outcome_summary,
  aes(x = outcome, y = median_absolute_scale_percent)
) +
  geom_col(fill = "#3C78A8") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median absolute ATE (% of outcome range)",
    title = "Typical forecast magnitude by outcome"
  ) +
  theme_minimal(base_size = 11)
```

# Primary median versus Nalanda mean sensitivity

The locked candidate uses Nalanda's native median at each level for robustness.
The sensitivity analysis reruns the same equal-weight hierarchy with arithmetic
means across completions, prompts, models, and families. The comparison below
shows whether that reducer choice materially changes the forecasts.

```{r aggregation-sensitivity, results='asis'}
if (mean_sensitivity_available) {
  median_method_gap <- median(
    aggregation_comparison$normalized_absolute_difference
  )
  max_method_gap <- max(
    aggregation_comparison$normalized_absolute_difference
  )

  cat(
    "- Median absolute method difference: **",
    percent(median_method_gap, accuracy = 0.01),
    " of the outcome scale**.\n",
    "- Maximum absolute method difference: **",
    percent(max_method_gap, accuracy = 0.01),
    " of the outcome scale**.\n\n",
    sep = ""
  )

  method_sensitive <- aggregation_comparison |>
    arrange(desc(normalized_absolute_difference)) |>
    slice_head(n = 12) |>
    mutate(
      mean_minus_median_scale_percent = 100 * normalized_difference
    ) |>
    select(
      condition,
      outcome,
      median_hierarchy_ate,
      mean_hierarchy_ate,
      mean_minus_median_scale_percent
    )

  print(kable_simple(
    method_sensitive,
    digits = 3,
    caption = "Cells most affected by median versus arithmetic-mean aggregation"
  ))
} else {
  cat(
    "> The Nalanda aggregation-comparison file is not available. Source ",
    "`scripts/03_build_submission.R` and regenerate this report.\n"
  )
}
```

# Prompt robustness

Prompt agreement tests whether fixed wording variants elicit similar forecasts
from the same model. Strong agreement is desirable for stability, but two
prompts can agree and still share the same bias.

```{r prompt-scatter, results='asis'}
if (n_prompts == 2L) {
  prompt_wide <- prompt_enriched |>
    select(
      condition, outcome, forecast_model, prompt_variant,
      prompt_ate_fraction
    ) |>
    pivot_wider(
      names_from = prompt_variant,
      values_from = prompt_ate_fraction
    )

  p1 <- prompt_names[[1L]]
  p2 <- prompt_names[[2L]]

  prompt_correlations <- prompt_wide |>
    group_by(forecast_model) |>
    summarize(
      pearson = cor(.data[[p1]], .data[[p2]], method = "pearson"),
      spearman = cor(.data[[p1]], .data[[p2]], method = "spearman"),
      .groups = "drop"
    )

  print(kable_simple(
    prompt_correlations,
    digits = 3,
    caption = "Prompt-variant agreement across the 208 cells"
  ))

  prompt_limit <- max(abs(c(prompt_wide[[p1]], prompt_wide[[p2]])))
  prompt_plot <- ggplot(
    prompt_wide,
    aes(x = .data[[p1]], y = .data[[p2]])
  ) +
    geom_abline(slope = 1, intercept = 0, color = "grey55") +
    geom_point(alpha = 0.65, color = "#2B6F8A") +
    facet_wrap(~forecast_model) +
    coord_equal(xlim = c(-prompt_limit, prompt_limit),
                ylim = c(-prompt_limit, prompt_limit)) +
    scale_x_continuous(labels = label_percent(accuracy = 0.1)) +
    scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
    labs(
      x = paste(p1, "ATE as % of scale"),
      y = paste(p2, "ATE as % of scale")
    ) +
    theme_minimal(base_size = 11)
  print(prompt_plot)
} else {
  cat("> Prompt-pair scatterplots require exactly two prompt variants.\n")
}
```

## Cells most sensitive to prompt wording

All models with complete full-run results are included below. A model can look
absent for a particular outcome when its median prompt gap is exactly zero:
a zero-height bar is invisible. The statistic is the median, across
interventions, of the absolute gap between the two fixed prompt variants for
each outcome and model. It measures prompt sensitivity, not forecast magnitude
or accuracy.

```{r prompt-sensitive-cells}
prompt_sensitive <- prompt_dispersion |>
  arrange(desc(prompt_range_fraction)) |>
  slice_head(n = 15) |>
  mutate(prompt_range_percent = 100 * prompt_range_fraction) |>
  select(
    forecast_model,
    condition,
    outcome,
    n_prompts,
    prompt_range_percent
  )

kable_simple(
  prompt_sensitive,
  digits = 2,
  caption = "Largest within-model prompt ranges"
)
```

```{r prompt-gap-plot}
prompt_by_outcome <- prompt_dispersion |>
  group_by(forecast_model, outcome) |>
  summarize(
    median_prompt_range_percent = 100 * median(prompt_range_fraction),
    .groups = "drop"
  ) |>
  mutate(outcome = factor(outcome, levels = outcome_order))

ggplot(
  prompt_by_outcome,
  aes(x = outcome, y = median_prompt_range_percent, fill = forecast_model)
) +
  geom_col(
    # coord_flip() otherwise makes the within-outcome bar order appear in the
    # opposite direction from the top-to-bottom legend order.
    position = position_dodge(width = 0.85, reverse = TRUE),
    width = 0.78,
    color = "white",
    linewidth = 0.15
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median prompt range (% of outcome scale)",
    fill = "Model",
    title = "Prompt sensitivity by outcome"
  ) +
  theme_minimal(base_size = 11)
```

```{r prompt-gap-interpretation, results='asis'}
most_prompt_sensitive <- prompt_by_outcome |>
  slice_max(median_prompt_range_percent, n = 1, with_ties = FALSE)

cat(
  '<div class="alert alert-warning"><strong>Largest current prompt signal:</strong> ',
  most_prompt_sensitive$forecast_model,
  " has the largest model–outcome median prompt gap: ",
  sprintf("%.1f%%", most_prompt_sensitive$median_prompt_range_percent),
  " of the outcome scale for <code>",
  as.character(most_prompt_sensitive$outcome),
  "</code>. This is a stability warning: the two fixed prompt formulations ",
  "produce materially different forecasts for that model. It does not, by ",
  "itself, reveal which forecast or model is more accurate.</div>\n",
  sep = ""
)
```

# Model agreement

```{r model-agreement, results='asis'}
if (n_models < 2L) {
  cat(
    "> Model-agreement statistics are intentionally withheld because only ",
    "one model is present. A 1 &times; 1 correlation of 1 is an identity, not ",
    "evidence of model consensus. This section will populate automatically ",
    "after additional models are run.\n"
  )
} else {
  pairwise_correlations <- read_csv(
    project_file(
      "outputs", "diagnostics", "model_correlations_pairwise.csv"
    ),
    show_col_types = FALSE
  )
  overall_agreement <- read_csv(
    project_file(
      "outputs", "diagnostics", "model_agreement_overall.csv"
    ),
    show_col_types = FALSE
  )
  model_disagreement <- read_csv(
    project_file(
      "outputs", "diagnostics", "model_disagreement_by_cell.csv"
    ),
    show_col_types = FALSE
  )

  print(kable_simple(
    overall_agreement,
    digits = 3,
    caption = "Overall scale-normalized model agreement"
  ))
  print(plot_model_agreement(
    pairwise_correlations,
    type = "heatmap",
    method = "spearman"
  ))

  model_spread <- model_disagreement |>
    arrange(desc(range_per_scale)) |>
    slice_head(n = 15) |>
    mutate(model_range_percent = 100 * range_per_scale) |>
    select(condition, outcome, model_range_percent)

  print(kable_simple(
    model_spread,
    digits = 2,
    caption = "Cells with the largest cross-model range"
  ))
}
```

# Interpretation checklist

Use this report to discuss the following questions with collaborators:

1. Do the largest positive and negative forecasts make substantive sense given
   the interventions and outcome directions?
2. Are any outcomes consistently assigned much larger effects than others
   after accounting for their native scales?
3. Which cells are sensitive to prompt wording, and are those disagreements
   large enough to justify another fixed prompt variant?
4. Once several models are present, do disagreements reflect magnitude,
   direction, or both?
5. Are any routing failures, missing cells, or manual repairs present? If yes,
   do not lock a submission until they are resolved and documented.

## What this report cannot establish

- It cannot identify the most accurate model without human benchmark outcomes.
- High prompt or model agreement does not imply calibration.
- Low agreement does not identify which forecast is correct.
- The report must not be used to remove a predeclared model merely because it
  disagrees with the others.

# Reproducibility information

```{r session-info}
cat("Generated from project root:", project_root, "\n\n")
cat("R version:", R.version.string, "\n")
cat("nalanda version:", as.character(utils::packageVersion("nalanda")), "\n")
cat("rmarkdown version:", as.character(utils::packageVersion("rmarkdown")), "\n")
```
