---
title: "Silicon Sample Benchmark Tier-3 Forecast"
subtitle: "Team 16 — primary three-model consensus"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    self_contained: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  fig.align = "center",
  dpi = 120
)

suppressPackageStartupMessages({
  library(nalanda)
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(scales)
  library(tidyr)
})

root_candidates <- unique(c(
  normalizePath(".", winslash = "/", mustWork = FALSE),
  normalizePath("..", winslash = "/", mustWork = FALSE)
))
is_project_root <- vapply(
  root_candidates,
  function(x) {
    file.exists(file.path(x, "inputs", "tier3_outcomes.csv")) &&
      file.exists(file.path(
        x, "outputs", "derived", "team_16_T3_candidate_median_unscaled.csv"
      ))
  },
  logical(1)
)
if (!any(is_project_root)) {
  stop("Could not locate the Silicon Benchmark project root.")
}

project_root <- root_candidates[which(is_project_root)[1]]
project_file <- function(...) file.path(project_root, ...)

candidate <- read_csv(
  project_file(
    "outputs", "derived", "team_16_T3_candidate_median_unscaled.csv"
  ),
  show_col_types = FALSE
)
outcomes <- read_csv(
  project_file("inputs", "tier3_outcomes.csv"),
  show_col_types = FALSE
)
interventions <- read_csv(
  project_file("inputs", "tier3_interventions.csv"),
  show_col_types = FALSE
)
models_config <- read_csv(
  project_file("config", "models.csv"),
  show_col_types = FALSE
)
primary_roster <- read_csv(
  project_file("outputs", "derived", "primary_model_roster.csv"),
  show_col_types = FALSE
)
forecast_stages <- read_csv(
  project_file(
    "outputs", "derived", "nalanda_forecast_aggregation_long.csv"
  ),
  show_col_types = FALSE
)
all_pairwise_correlations <- read_csv(
  project_file("outputs", "diagnostics", "model_correlations_pairwise.csv"),
  show_col_types = FALSE
)
primary_prompt_level <- forecast_stages |>
  filter(
    aggregation_level == "prompt",
    model_id %in% primary_roster$model_id
  ) |>
  transmute(
    condition,
    outcome,
    forecast_model = model_id,
    prompt_variant = prompt_id,
    prompt_ate = estimate
  )
primary_model_level <- forecast_stages |>
  filter(
    aggregation_level == "model",
    model_id %in% primary_roster$model_id
  ) |>
  transmute(
    condition,
    outcome,
    forecast_model = model_id,
    model_ate = estimate
  )
primary_prompt_dispersion <- read_csv(
  project_file(
    "outputs", "diagnostics", "prompt_disagreement_by_cell.csv"
  ),
  show_col_types = FALSE
) |>
  filter(model_id %in% primary_roster$model_id) |>
  transmute(
    condition,
    outcome,
    forecast_model = model_id,
    n_prompts = n_contributors,
    prompt_range_fraction = range_per_scale
  )

primary_models <- models_config |>
  filter(model_id %in% primary_roster$model_id) |>
  arrange(family, model_id) |>
  select(model_id, family, temperature, n_completions, output_mode)

model_labels <- c(
  "claude-opus-4-6" = "Claude Opus 4.6",
  "gemini-3.1-pro-preview" = "Gemini 3.1 Pro Preview",
  "gpt-5-mini" = "GPT-5 mini"
)
primary_correlations <- all_pairwise_correlations |>
  filter(
    model_a %in% primary_roster$model_id,
    model_b %in% primary_roster$model_id
  ) |>
  mutate(
    model_pair = paste(
      unname(model_labels[model_a]),
      unname(model_labels[model_b]),
      sep = " vs. "
    ),
    statistic = recode(method, pearson = "Pearson", spearman = "Spearman")
  ) |>
  select(model_pair, statistic, correlation, n_units)

primary_prompt_enriched <- primary_prompt_level |>
  left_join(
    transmute(outcomes, outcome, scale_width = native_max - native_min),
    by = "outcome"
  ) |>
  mutate(prompt_ate_fraction = prompt_ate / scale_width)

primary_model_disagreement <- primary_model_level |>
  left_join(
    transmute(outcomes, outcome, scale_width = native_max - native_min),
    by = "outcome"
  ) |>
  group_by(condition, outcome) |>
  summarize(
    model_range_percent = 100 * (max(model_ate) - min(model_ate)) /
      first(scale_width),
    .groups = "drop"
  )

prompt_names <- sort(unique(primary_prompt_level$prompt_variant))

outcome_order <- outcomes$outcome
condition_order <- sort(unique(interventions$condition))
candidate_enriched <- candidate |>
  left_join(
    select(outcomes, outcome, native_min, native_max, definition),
    by = "outcome"
  ) |>
  mutate(
    scale_width = native_max - native_min,
    ate_fraction = ate / scale_width,
    outcome = factor(outcome, levels = outcome_order),
    condition = factor(condition, levels = rev(condition_order))
  )
fill_limit <- max(abs(candidate_enriched$ate_fraction), na.rm = TRUE)

stopifnot(
  nrow(candidate) == 208L,
  n_distinct(candidate$condition) == 16L,
  n_distinct(candidate$outcome) == 13L,
  !anyNA(candidate$ate),
  nrow(primary_models) == 3L,
  n_distinct(primary_models$family) == 3L,
  nrow(primary_correlations) == 6L,
  length(prompt_names) == 2L
)
```

# Submission summary

This Tier-3 entry predicts the average treatment effect (ATE) of each of 16
interventions on each of 13 outcomes, for a complete set of **208 point
forecasts**. It uses one preselected language model from each of three developer
families and gives the families equal weight.

- Entry: **team_16 primary Tier 3**
- Forecasting approach: **direct structured LLM forecasts**
- Primary ensemble: **three developer families, equally weighted**
- Aggregation: **median within prompts, models, and families**
- Post-hoc adjustment: **none in the primary candidate**
- Coverage: **16 interventions × 13 outcomes; no missing values**

# Primary model specification

```{r model-specification}
knitr::kable(
  primary_models,
  caption = "Preselected contributors to the primary consensus"
)
```

Each model received two fixed direct-forecast prompt formulations for every
intervention. GPT-5-mini used five completions per formulation because its
gateway temperature is fixed at 1; the other two models used one completion per
formulation. Completions were first combined within each prompt, then prompts
within each model. Because there is one selected model per family, the final
family-balanced consensus gives Anthropic, Google, and OpenAI equal weight.

# Understanding the forecasts

<div class="alert alert-info">
<strong>Average treatment effect (ATE)</strong><br>
An ATE is the forecast average outcome in an intervention group minus the
forecast average in the neutral control group. A positive ATE means the
intervention group is expected to score higher; a negative ATE means it is
expected to score lower. The heatmap divides every ATE by that outcome's full
scale width so outcomes measured in dollars, proportions, and 0–100 ratings can
be displayed together. For example, a $1 donation ATE is 10% of the $0–$10
donation scale, while a five-point ATE on a 0–100 rating is 5% of its scale.
</div>

```{r primary-heatmap, fig.height=8}
ggplot(candidate_enriched, aes(x = outcome, y = condition, fill = ate_fraction)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-fill_limit, fill_limit),
    labels = label_percent(accuracy = 0.1),
    name = "ATE as %\nof scale"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Primary three-family consensus forecasts"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

The pale or saturated color of a tile reflects the magnitude of its forecast
relative to that outcome's possible range. It does not represent uncertainty or
accuracy; human benchmark outcomes were unavailable when these predictions were
generated and locked.

# Forecast magnitude by outcome

```{r outcome-calibration}
outcome_summary <- candidate_enriched |>
  group_by(outcome) |>
  summarize(
    median_absolute_scale_percent = 100 * median(abs(ate_fraction)),
    maximum_absolute_scale_percent = 100 * max(abs(ate_fraction)),
    .groups = "drop"
  ) |>
  mutate(outcome = reorder(outcome, median_absolute_scale_percent))

ggplot(
  outcome_summary,
  aes(x = outcome, y = median_absolute_scale_percent)
) +
  geom_col(fill = "#3C78A8") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median absolute ATE (% of outcome range)",
    title = "Typical primary-forecast magnitude by outcome"
  ) +
  theme_minimal(base_size = 11)
```

# Prompt robustness

Prompt agreement tests whether the two fixed wording variants elicit similar
forecasts from the same primary model. Strong agreement is desirable for
stability, but two prompts can agree and still share the same bias.

```{r primary-prompt-scatter}
primary_prompt_wide <- primary_prompt_enriched |>
  select(
    condition, outcome, forecast_model, prompt_variant,
    prompt_ate_fraction
  ) |>
  pivot_wider(
    names_from = prompt_variant,
    values_from = prompt_ate_fraction
  )

p1 <- prompt_names[[1L]]
p2 <- prompt_names[[2L]]

primary_prompt_correlations <- primary_prompt_wide |>
  group_by(forecast_model) |>
  summarize(
    Pearson = cor(.data[[p1]], .data[[p2]], method = "pearson"),
    Spearman = cor(.data[[p1]], .data[[p2]], method = "spearman"),
    .groups = "drop"
  ) |>
  mutate(
    Model = unname(model_labels[forecast_model]),
    .before = 1
  ) |>
  select(Model, Pearson, Spearman)

knitr::kable(
  primary_prompt_correlations,
  format = "html",
  digits = 3,
  caption = "Prompt-variant agreement across the 208 cells"
)

prompt_limit <- max(abs(c(
  primary_prompt_wide[[p1]], primary_prompt_wide[[p2]]
)))

ggplot(
  primary_prompt_wide |>
    mutate(Model = unname(model_labels[forecast_model])),
  aes(x = .data[[p1]], y = .data[[p2]])
) +
  geom_abline(slope = 1, intercept = 0, color = "grey55") +
  geom_point(alpha = 0.65, color = "#2B6F8A") +
  facet_wrap(~Model) +
  coord_equal(
    xlim = c(-prompt_limit, prompt_limit),
    ylim = c(-prompt_limit, prompt_limit)
  ) +
  scale_x_continuous(labels = label_percent(accuracy = 0.1)) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(
    x = paste(p1, "ATE as % of scale"),
    y = paste(p2, "ATE as % of scale")
  ) +
  theme_minimal(base_size = 11)
```

## Cells most sensitive to prompt wording

The table identifies individual primary-model cells with the largest absolute
gap between the two prompt variants. The figure then summarizes, for each
outcome and model, the median gap across interventions. These are stability
diagnostics, not measures of forecast magnitude or accuracy.

```{r primary-prompt-sensitive-cells}
primary_prompt_sensitive <- primary_prompt_dispersion |>
  arrange(desc(prompt_range_fraction)) |>
  slice_head(n = 15) |>
  mutate(
    Model = unname(model_labels[forecast_model]),
    `Prompt range (% of scale)` = 100 * prompt_range_fraction
  ) |>
  select(
    Model,
    Intervention = condition,
    Outcome = outcome,
    `Prompt range (% of scale)`
  )

knitr::kable(
  primary_prompt_sensitive,
  format = "html",
  digits = 2,
  caption = "Largest within-model prompt ranges among primary models"
)
```

```{r primary-prompt-gap-plot}
primary_prompt_by_outcome <- primary_prompt_dispersion |>
  group_by(forecast_model, outcome) |>
  summarize(
    median_prompt_range_percent = 100 * median(prompt_range_fraction),
    .groups = "drop"
  ) |>
  mutate(
    Model = unname(model_labels[forecast_model]),
    outcome = factor(outcome, levels = outcome_order)
  )

ggplot(
  primary_prompt_by_outcome,
  aes(x = outcome, y = median_prompt_range_percent, fill = Model)
) +
  geom_col(
    position = position_dodge(width = 0.85, reverse = TRUE),
    width = 0.78,
    color = "white",
    linewidth = 0.15
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Median prompt range (% of outcome scale)",
    fill = "Model",
    title = "Prompt sensitivity by outcome among primary models"
  ) +
  theme_minimal(base_size = 11)
```

# Agreement among the primary models

<div class="alert alert-info">
<strong>What model agreement measures</strong><br>
Agreement asks whether the three primary models assign similar relative effect
patterns across the same 208 intervention–outcome cells. Pearson correlation
emphasizes similarity in forecast magnitudes, while Spearman correlation
emphasizes similarity in rank ordering. High agreement supports the claim that
the consensus is not driven by a single provider family, but it is not evidence
that the forecasts are accurate; accuracy can be assessed only after comparison
with the locked human benchmark results.
</div>

```{r primary-model-agreement}
agreement_table <- primary_correlations |>
  mutate(correlation = round(correlation, 3)) |>
  tidyr::pivot_wider(
    names_from = statistic,
    values_from = correlation
  ) |>
  arrange(model_pair) |>
  select(
    `Model pair` = model_pair,
    Pearson,
    Spearman,
    `Forecast cells` = n_units
  )

knitr::kable(
  agreement_table,
  format = "html",
  escape = TRUE,
  caption = "Pairwise agreement among the three primary model forecasts"
)

plot_model_agreement(
  all_pairwise_correlations |>
    filter(
      model_a %in% primary_roster$model_id,
      model_b %in% primary_roster$model_id
    ),
  type = "heatmap",
  method = "spearman"
)
```

Across the three model pairs, Pearson correlations range from
`r sprintf("%.3f", min(primary_correlations$correlation[primary_correlations$statistic == "Pearson"]))`
to
`r sprintf("%.3f", max(primary_correlations$correlation[primary_correlations$statistic == "Pearson"]))`,
and Spearman correlations range from
`r sprintf("%.3f", min(primary_correlations$correlation[primary_correlations$statistic == "Spearman"]))`
to
`r sprintf("%.3f", max(primary_correlations$correlation[primary_correlations$statistic == "Spearman"]))`.
The full all-model agreement and prompt-sensitivity analyses remain in the
separate exploratory diagnostics report.

```{r primary-model-sensitive-cells}
primary_model_spread <- primary_model_disagreement |>
  arrange(desc(model_range_percent)) |>
  slice_head(n = 15) |>
  select(
    Intervention = condition,
    Outcome = outcome,
    `Model range (% of scale)` = model_range_percent
  )

knitr::kable(
  primary_model_spread,
  format = "html",
  digits = 2,
  caption = "Cells with the largest range across the three primary models"
)
```

# Reproducibility note

The prediction pipeline retains the fixed prompt texts, model routing and
generation settings, structured completion records with row-level provenance,
hierarchical aggregation outputs, and the final Tier-3 file. The definitive
method disclosure, blinding attestation, model-call counts, costs, and archive
locations are recorded in the benchmark's required `registration.md` and
`metadata.json`. This report is a supplementary human-readable summary; the
locked prediction CSV and completed registration are the scored submission
materials.

```{r session-information}
sessionInfo()
```
